Type Evolution in Interactive Programming Systems

### 2024-05-30 paper.tex
tweaks to my stuff

\challenge{Extract Entity}
Acme needs to evolve the above schema and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
one with orders that links to one with customers.  We refer to this as \emph{Extract Entity}
operation, because it introduces a new \emph{entity}, a referenceable holder of mutable attributes.
The operation is needed when we realize that some attributes of one type of entity should belong to
a  distinct type of entity that will be associated with the first one.

@@

The \textit{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers,
and 3) reference these identifiers from the orders. Our example took the simple approach of
merging entities whose attributes are all equal. But that might not be correct in all cases.
What if there was a typo in one of the addresses that made it look different from other instances
of the customer? To repair that error we need to merge the falsely distinct customers into one
and fix any references from orders. We call this operation \textit{Merge Entities} -- note that
it is not a schema change but a data change, yet nevertheless not commonly supported in data APIs.

Entity evolution can happen through operations such as those listed in
Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
change in requirements that triggered an evolution might change back again. Inverse
is also required in research on \emph{divergence control}~\cite{Foster2007, herrmann17, chillon22}.
The inverse of \textit{Extract Entity} is \textit{Absorb Entity}. It raises the question of how
to handle \textit{one-to-many} relationships. For example should absorbing orders into customers
do a relational join, returning us back to the starting point, or should it nest orders within
customers, as a NoSQL database might do?

The inverse of \textit{Merge Entities} is \textit{Split Entity}, but that can be awkward.
Imagine that we didn't have addresses for customers to disambiguate them with and only knew
their non-unique names. Then \textit{Extract Entity} will conflate them and discard the
information needed to properly invert the operation. Should this information be retained?
One could object that we shouldn't have taken orders for people we can't uniquely identify
yet. And even if that was necessary it means the customers really aren't distinct entities
and shouldn't have been extracted. These are arguments that we shouldn't need \textit{Split
Entity}, contradicting the principle that evolution operations should be invertible.
It is interesting to observe that evolution in nature can also fail to invert, leaving
behind vestigial features. We consider this to be an open issue.

@@

\noindent
After applying the \textit{Extract Entity} operation as described earlier, the code should
be rewritten into the following query that joins the two linked tables:

@@

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code. \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accomodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}. BiDEL appears capable of handling the basic requirements of our \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} challenges, though \textit{Split/Merge Entity} is less clear. It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.

@@

Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases. While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}. \citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON. \citet{scherzinger13} define a set of operators like the relational SMOs discussed above. \citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} but not \textit{Split/Merge Entity} nor divergence. None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.

@@

\paragraph{Model Driven Engineering.}
Model Driven Engineering must also handle evolution. Unlike the textual artifacts involved in other domains, models typically assign unique indentifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
Models are often themselves modeled by a metamodel, which lifts the evolution problem to the metalevel: models must be migrated through the evolution of their metamodel, analogously to changing the grammar of a programming language.~\cite{Herrmannsdoerfer11}. \citet{Cicchetti11} study metamodel evolution in the context of divergent changes.
\citet{vermolen11} define a DSL for evolving a data model by generating SQL migrations on the backing database. They consider the evolution of OO-style subclass hierarchies which goes beyond most other work. Their running example includes both our Extract Entity and Multiplicty Change examples, and would make for a good follow-on problem. However they do not address code/query rewriting.

>>>>

\challenge{Extract Entity}
Acme needs to evolve the above schema and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
one with orders that links to one with customers.  We refer to this as the \emph{Extract Entity}
operation, because it introduces a new \emph{entity}, a referenceable holder of mutable attributes.
The operation is needed when we realize that some attributes of one type of entity should belong to
a  distinct type of entity that will be associated with the first one.

@@

The \emph{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers,
and 3) reference these identifiers from the orders. Our example took the simple approach of
merging entities whose attributes are all equal. But that might not be correct in all cases.
What if there was a typo in one of the addresses that made it look different from other instances
of the customer? To repair that error we need to merge the falsely distinct customers into one
and fix any references from orders. We call this operation \emph{Merge Entities} -- note that
it is not a schema change but a data change, yet nevertheless not commonly supported in data APIs.

Entity evolution can happen through operations such as those listed in
Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
change in requirements that triggered an evolution might change back again.
Invertibility features in other work on schema evolution and \emph{bidirectional transformations}~\cite{Foster2007, herrmann17, chillon22}.
The inverse of \emph{Extract Entity} is \emph{Absorb Entity}. It raises the question of how
to handle \emph{one-to-many} relationships. For example should absorbing orders into customers
do a relational join, returning us back to the starting point, or should it nest orders within
customers, as a NoSQL database might do?

The inverse of \emph{Merge Entities} is \emph{Split Entity}, but that can be awkward.
Imagine that we didn't have addresses for customers to disambiguate them with and only knew
their non-unique names. Then \emph{Extract Entity} will conflate them and discard the
information needed to properly invert the operation. Should this information be retained?
One could object that we shouldn't have taken orders for people we can't uniquely identify
yet. And even if that was necessary it means the customers really aren't distinct entities
and shouldn't have been extracted. These are arguments that we shouldn't need \textit{Split
Entity}, contradicting the principle that evolution operations should be invertible.
It is interesting to observe that evolution in nature can also fail to invert, leaving
behind vestigial features. We consider this to be an open issue.

@@

\noindent
After applying the \emph{Extract Entity} operation as described earlier, the code should
be rewritten into the following query that joins the two linked tables:

@@

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code. \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accomodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}. BiDEL appears capable of handling the basic requirements of our \emph{Extract/Absorb Entity} and \emph{Multiplicity Change} challenges, though \emph{Split/Merge Entity} is less clear. It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.

@@

Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases. While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}. \citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON. \citet{scherzinger13} define a set of operators like the relational SMOs discussed above. \citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling \textit{Extract/Absorb Entity} and \emph{Multiplicity Change} but not \emph{Split/Merge Entity} nor divergence. None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.

@@

\paragraph{Model Driven Engineering.}
Model Driven Engineering must also handle evolution. Unlike the textual artifacts involved in other domains, models typically assign unique indentifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
Models are often themselves modeled by a metamodel, which lifts the evolution problem to the metalevel: models must be migrated through the evolution of their metamodel, analogously to changing the grammar of a programming language.~\cite{Herrmannsdoerfer11}. \citet{Cicchetti11} study metamodel evolution in the context of divergent changes.
\citet{vermolen11} define a DSL for evolving a data model by generating SQL migrations on the backing database. They consider the evolution of OO-style subclass hierarchies which goes beyond most other work. Their running example includes both our \emph{Extract Entity} and \emph{Multiplicty Change} examples, and would make for a good follow-on problem. However they do not address code/query rewriting.
